Approximation as Differentiation

### 2025-07-08 main/models.tex
semi-lattice -> semilattice.

\begin{definition}
  \begin{enumerate}
  \item The category $\LatGal$ has bounded lattices as objects and
    Galois connections as morphisms, with the right adjunct going in
    the ``forward'' direction.
  \item The category $\MeetSLat$ has meet-semi-lattices with top as
    objects and monotone finite meet preserving functions as
    morphisms.
  \item The category $\JoinSLat$ has join-semi-lattices with bottom as
    objects and monotone finite join preserving functions as
    morphisms.
  \end{enumerate}
\end{definition}

@@

\begin{example}
  The following categories are $\CMon$-enriched and have finite
  products, and hence biproducts:
  \begin{enumerate}
  \item In $\FinVect$, morphisms are linear maps and so can be added
    and have a zero map. Finite products are given by cartesian
    products of the underlying sets, with the vector operations
    defined pointwise.
  \item In $\LatGal$, right adjoints are summed using meets and left
    adjoints are summed using joins. The zero maps are given by the
    constantly $\top$ and constantly $\bot$ functions
    respectively. Products are given by the cartesian product of the
    underlying set.
  \item The $\LatGal$ example splits into two separate categories
    $\MeetSLat$ and $\JoinSLat$ of meet semi-lattices and join
    semi-lattices respectively, with their respective homomorphisms as
    the morphisms. As for $\LatGal$, these categories are
    $\CMon$-enriched and have finite products, and so finite
    biproducts.
  \end{enumerate}
\end{example}

@@

We will avoid the need for infinitary operations by separating the
forward and backward parts of the Galois connections to act
independently by moving to the product category
$\MeetSLat \times \JoinSLat^\op$. Objects in this category consist of
\emph{separate} meet- and join-semi-lattices and potentially unrelated
forward meet-preserving and backward join-preserving maps. We first
check that this category satisfies the hypotheses of
\thmref{fam-closed}:

@@

  $\MeetSLat$ has all products, indeed all limits, because it is the
  category of algebras for a Lawvere theory. Similarly, $\JoinSLat$
  has all coproducts, indeed all colimits, for the same reason. Note
  that these are very different constructions: elements of a product
  of meet-semi-lattices consist of (possibly infinite) tuples of
  elements, while elements of a coproduct of join-semilattices consist
  of \emph{finite} formal joins of elements quotiented by the
  join-semi-lattice equations. Since $\JoinSLat$ has all coproducts,
  $\JoinSLat^\op$ has all products, and so
  $\MeetSLat \times \JoinSLat^\op$ has all products, as required.
\end{proof}

@@

With this embedding functor, we can see that the interpretation of
first-order types will be the same up to isormorphism in
$\Fam(\LatGal)$ and $\Fam(\MeetSLat \times \JoinSLat^\op)$, as long as
we interpret the base types as objects in $\Fam(\LatGal)$. At
higher-order, however, the meet-semi-lattice and join-semi-lattice
sides of the interpretation will diverge, and it is no longer clear
that the interpretation of programs using higher-order functions
internally will result in Galois connections. In \secref{definability}
we will see that every program with first-order type (even if it uses
higher-order functions internally) does in fact have an interpretation
definable in $\Fam(\LatGal)$.

>>>>

\begin{definition}
  \begin{enumerate}
  \item The category $\LatGal$ has bounded lattices as objects and
    Galois connections as morphisms, with the right adjunct going in
    the ``forward'' direction.
  \item The category $\MeetSLat$ has meet-semilattices with top as
    objects and monotone finite meet preserving functions as
    morphisms.
  \item The category $\JoinSLat$ has join-semilattices with bottom as
    objects and monotone finite join preserving functions as
    morphisms.
  \end{enumerate}
\end{definition}

@@

\begin{example}
  The following categories are $\CMon$-enriched and have finite
  products, and hence biproducts:
  \begin{enumerate}
  \item In $\FinVect$, morphisms are linear maps and so can be added
    and have a zero map. Finite products are given by cartesian
    products of the underlying sets, with the vector operations
    defined pointwise.
  \item In $\LatGal$, right adjoints are summed using meets and left
    adjoints are summed using joins. The zero maps are given by the
    constantly $\top$ and constantly $\bot$ functions
    respectively. Products are given by the cartesian product of the
    underlying set.
  \item The $\LatGal$ example splits into two separate categories
    $\MeetSLat$ and $\JoinSLat$ of meet semilattices and join
    semilattices respectively, with their respective homomorphisms as
    the morphisms. As for $\LatGal$, these categories are
    $\CMon$-enriched and have finite products, and so finite
    biproducts.
  \end{enumerate}
\end{example}

@@

We will avoid the need for infinitary operations by separating the
forward and backward parts of the Galois connections to act
independently by moving to the product category
$\MeetSLat \times \JoinSLat^\op$. Objects in this category consist of
\emph{separate} meet- and join-semilattices and potentially unrelated
forward meet-preserving and backward join-preserving maps. We first
check that this category satisfies the hypotheses of
\thmref{fam-closed}:

@@

  $\MeetSLat$ has all products, indeed all limits, because it is the
  category of algebras for a Lawvere theory. Similarly, $\JoinSLat$
  has all coproducts, indeed all colimits, for the same reason. Note
  that these are very different constructions: elements of a product
  of meet-semilattices consist of (possibly infinite) tuples of
  elements, while elements of a coproduct of join-semilattices consist
  of \emph{finite} formal joins of elements quotiented by the
  join-semilattice equations. Since $\JoinSLat$ has all coproducts,
  $\JoinSLat^\op$ has all products, and so
  $\MeetSLat \times \JoinSLat^\op$ has all products, as required.
\end{proof}

@@

With this embedding functor, we can see that the interpretation of
first-order types will be the same up to isormorphism in
$\Fam(\LatGal)$ and $\Fam(\MeetSLat \times \JoinSLat^\op)$, as long as
we interpret the base types as objects in $\Fam(\LatGal)$. At
higher-order, however, the meet-semilattice and join-semilattice
sides of the interpretation will diverge, and it is no longer clear
that the interpretation of programs using higher-order functions
internally will result in Galois connections. In \secref{definability}
we will see that every program with first-order type (even if it uses
higher-order functions internally) does in fact have an interpretation
definable in $\Fam(\LatGal)$.
